---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  output_hook(gsub("[ \\t]+(?=\\n|$)", "", x, perl = TRUE), options)
})
```

# lmmix

`lmmix` fits Gaussian linear mixed models by explicitly optimizing the ML or
REML criterion. Its main use case is a model that contains both random effects
and a structured residual covariance.

The package estimates the covariance parameters, fixed effects and empirical
BLUPs in one fitted object. It also provides Satterthwaite and Kenward-Roger
inference, type III tests, likelihood-ratio comparisons, estimated marginal
means, pairwise contrasts, standard model methods, `broom` methods and
`emmeans` interoperability.

The package is under active development. The table below summarizes the
implemented functionality and the intended model boundary.

## Installation

Install the development version directly from the public GitHub repository:

```{r installation, eval=FALSE}
install.packages("devtools")
devtools::install_github("AurelienNicosiaULaval/lmmix")
```

## Vignettes

The package contains five complementary vignettes:

* [Using lmmix](articles/lmmix.html)
  ([source](vignettes/lmmix.Rmd))
  presents the fitting workflow and the main methods;
* [Estimation and inference in lmmix](articles/theory-reml.html)
  ([source](vignettes/theory-reml.Rmd))
  gives the mathematical formulation, covariance parameterizations, ML and
  REML criteria, and Satterthwaite and Kenward-Roger inference;
* [Covariance intervals and boundary-aware model comparisons](articles/boundary-inference.html)
  ([source](vignettes/boundary-inference.Rmd))
  explains bound-respecting covariance intervals and parametric-bootstrap
  likelihood-ratio tests;
* [Validation against independent R implementations and PROC MIXED](articles/validation.html)
  ([source](vignettes/validation.Rmd))
  documents the independent comparisons, numerical tolerances, and validation
  boundaries;
* [Choosing lmmix and related R packages](articles/comparison.html)
  ([source](vignettes/comparison.Rmd))
  compares the implemented model and inference scope with `lme4`, `lmerTest`,
  `nlme`, and `mmrm`.

They are also available from an installed package:

```{r open-vignettes, eval=FALSE}
vignette("lmmix", package = "lmmix")
vignette("theory-reml", package = "lmmix")
vignette("boundary-inference", package = "lmmix")
vignette("validation", package = "lmmix")
vignette("comparison", package = "lmmix")
```

## Current capability matrix

| Component | Current behavior |
|:--|:--|
| Response | Univariate continuous Gaussian response |
| Estimation | Explicit profiled ML and REML optimization |
| Random effects | No random effect, one grouping formula, or independent crossed or nested formulas with correlated slopes within each term |
| Residual covariance | `id`, `cs`, `ar1`, full `toep`, banded `toep(k)`, and `un` |
| Combined covariance | Random effects and any supported residual structure in the same model |
| Denominator degrees of freedom | Satterthwaite, Kenward-Roger, or residual |
| Fixed-effect tests | Coefficient tests, type III tests, and nested-model likelihood-ratio comparisons |
| Covariance inference | Bound-respecting Wald intervals and parametric-bootstrap likelihood-ratio tests |
| Marginal means | Equal-weight estimated marginal means and automatic pairwise contrasts |
| Multiplicity | Any `stats::p.adjust()` method and simultaneous Bonferroni confidence intervals |
| Missing data | `na.omit`, `na.exclude`, or `na.fail` |
| Interoperability | `emmeans`, `tidy()`, `glance()`, and `augment()` |
| Diagnostic plots | `plot.lmm()` returns `ggplot2` residual, Q-Q, and fitted-value diagnostics |
| Standard extensions | Known precision weights, offsets, custom contrasts, simulation, updating, and prediction intervals |
| Model boundary | Gaussian responses; blockwise likelihood when independent components exist, with dense fitted covariance and inference matrices |

The [development roadmap](https://aureliennicosiaulaval.github.io/lmmix/ROADMAP.html)
centralizes the version history, current limitations, validation boundaries,
and planned development priorities.

## Quick start: random center effect and AR(1) residuals

The included `multicentre` data are a three-center repeated-measures example
with three treatments and three measurement occasions. The data contain 153
rows, including 28 missing responses. Their source is Milliken and Johnson
(2009, Section 28.3).

```{r combined-fit}
library(lmmix)

fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Drug:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

fit
```

By default, `lmm()` omits rows that are incomplete for variables required by
the fixed, random and repeated formulas. The fitted object therefore records
125 observations for this example. Use `na.action = na.exclude` to restore
omitted positions in fitted values, residuals and augmented data.

```{r combined-inference}
nobs(fit)
anova(fit)
VarCorr(fit)
```

Estimated marginal means and their pairwise differences are generated from
the fixed-effects reference grid. No contrast rows need to be written by hand.

```{r combined-lsmeans}
lsmeans(fit, ~Drug)
lsmeans(fit, pairwise ~ Drug, adjust = "holm")
```

The convergence code and Hessian diagnostic should be checked before
interpreting inference.

```{r convergence}
fit$convergence[c(
  "code",
  "optimizer",
  "hessian_positive_definite"
)]
```

## Three supported model families

The same function covers three model families.

```{r model-families, eval=FALSE}
# Marginal repeated-measures model
marginal_fit <- lmm(
  data,
  response ~ treatment * time,
  repeated = ~ time | subject,
  structure = "un"
)

# Random-effects model with independent residuals
random_fit <- lmm(
  data,
  response ~ treatment + time,
  random = ~ 1 + time | subject,
  structure = "id"
)

# Crossed random intercepts
crossed_fit <- lmm(
  data,
  response ~ treatment + time,
  random = list(
    center = ~ 1 | center,
    subject = ~ 1 | subject
  )
)

# Combined random-effect and correlated-residual model
combined_fit <- lmm(
  data,
  response ~ treatment * time,
  random = ~ 1 | center,
  repeated = ~ time | center:subject,
  structure = "ar1"
)
```

For `cs`, `ar1`, `toep`, or `un`, `repeated` must identify one ordering
variable and one grouping expression. Each repeated-measures group may contain
at most one observation at a given ordering value.

## Working with a fitted model

Standard methods return fixed effects, random effects, covariance components,
predictions, residuals and information criteria.

```{r model-methods}
fixef(fit)
head(ranef(fit)[[1L]])
head(residuals(fit))
head(predict(fit))
```

The `generics` methods return tibbles suitable for downstream analysis.
Printed tables use readable headings without dots, including `Std Error`,
`p value`, `Conf Low`, and `Conf High`. The underlying objects retain the
standard `broom` names required for programmatic use.

```{r broom-methods}
generics::tidy(fit)
generics::glance(fit)
head(generics::augment(fit))
```

The S3 `plot()` method returns a `ggplot2` object and supports residual,
normal Q-Q, and observed-versus-fitted diagnostics.

```{r diagnostic-plot, fig.alt="Standardized residuals plotted against fitted values"}
plot(fit, which = "residuals")
```

When `emmeans` is installed, the package supplies the fixed-effects basis and
model-specific denominator degrees of freedom.

```{r emmeans-example, eval=requireNamespace("emmeans", quietly = TRUE)}
emmeans::emmeans(fit, ~Drug)
```

`predict()` can return standard errors, confidence intervals, or prediction
intervals. `simulate()` draws marginal Gaussian responses from the fitted
model, and `update()` refits a modified specification.

```{r standard-extensions, eval=FALSE}
predict(fit, se.fit = TRUE, re.form = NA)
predict(fit, interval = "prediction", re.form = NA)
simulate(fit, nsim = 20, seed = 2026)
update(fit, . ~ . - Drug:Time)
```

Prediction standard errors condition on any included empirical random effects.
Prediction intervals add the fitted residual variance; they do not add BLUP
estimation uncertainty.

Kenward-Roger inference is available for REML fits. Nested models with
different fixed effects are automatically refitted with ML for a
likelihood-ratio comparison.

```{r extended-inference, eval=FALSE}
kr_fit <- lmm(
  data,
  response ~ treatment * time,
  random = ~ 1 | subject,
  ddf = "kenward-roger"
)

anova(reduced_fit, full_fit)
```

## Validation

The test suite separates numerical agreement from general software tests.

| Model component | External comparison | Quantities checked |
|:--|:--|:--|
| Random intercept and random slope | `nlme::lme()` | Fixed effects, covariance parameters, log-likelihood |
| Marginal CS and AR(1) | `nlme::gls()` | Fixed effects, covariance parameters, log-likelihood |
| Marginal AR(1) | `mmrm::mmrm()` | Fixed effects, residual covariance, log-likelihood |
| Satterthwaite inference | `lmerTest::lmer()` | Coefficient degrees of freedom and type III tests |
| Kenward-Roger inference | `lmerTest` and `mmrm` | Adjusted covariance and denominator degrees of freedom |
| Crossed random effects | `lme4::lmer()` | Fixed effects, covariance parameters and log-likelihood |
| Nested-model comparisons | `lme4::anova()` | Likelihood-ratio statistics and p-values |
| Official split-plot example 79.1 | Published PROC MIXED output | Variance components, REML criterion and type III tests |
| Official repeated-measures example 79.2 | Published PROC MIXED output | UN and CS covariance, ML criterion, fixed effects and type III statistics |
| Official random-coefficients example 79.5 | Published PROC MIXED output | Random covariance, REML criterion, fixed effects, BLUPs and type III test |
| Official line-source example 79.6 | Published PROC MIXED output | Three random variances, banded `TOEP(4)` covariance and REML criterion |
| Combined random center and AR(1) | Stored PROC MIXED targets | Covariance parameters, fixed effects, type III tests and marginal means |

The package also tests positive definiteness and convergence for every
supported residual structure in a combined model. See `VALIDATION.md` and the
validation vignette for model specifications, tolerances and limitations.

## Model scope and remaining boundaries

The complete and maintained list is available in the
[development roadmap](https://aureliennicosiaulaval.github.io/lmmix/ROADMAP.html).

Version `0.4.0` fits univariate Gaussian models. It is not a generalized
mixed-model engine. The likelihood detects independent connected components
and factors their covariance blocks separately. The fitted object and some
inference calculations still retain dense marginal matrices. Fully crossed
designs may also form one large connected component, so the package does not
yet make a large-scale sparse-computation claim.

Rows with missing responses or model covariates are handled according to
`na.action`; the package does not impute missing covariates. Conditional
prediction rejects unseen random-effect groups by default. Set
`allow.new.levels = TRUE` to request the population-level prediction, for
which the unseen random contribution is zero.

## Theoretical sources

The REML implementation follows Patterson and Thompson (1971), Harville
(1977), and LaMotte (2007). Satterthwaite inference follows Satterthwaite
(1946), with multi-degree-of-freedom tests based on Fai and Cornelius (1996).
Small-sample covariance adjustment follows Kenward and Roger (1997).
Estimated marginal means follow Searle, Speed, and Milliken (1980). Full
citations and DOI links are included in the package help and the
[theory vignette](articles/theory-reml.html). The
[comparison vignette](articles/comparison.html) cites the primary software
papers and official package documentation used to delimit overlapping scope.

## Package citation

The installed package provides its current academic citation directly:

```r
citation("lmmix")
```
