---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# lmmix

`lmmix` fits Gaussian linear mixed models by explicitly optimizing the ML or
REML criterion for

\[
V = ZGZ^\mathsf{T} + R.
\]

The package is designed for models that combine random effects with correlated
residuals. It provides Satterthwaite denominator degrees of freedom, type III
tests, estimated marginal means, pairwise contrasts, broom methods, and
`emmeans` interoperability.

The package name is `lmmix`; the main fitting function is `lmm()`.

## Installation

Install the package from a local checkout:

```{r installation, eval=FALSE}
devtools::install("path/to/lmmix")
```

## Combined random-effect and residual-correlation model

The included `multicentre` data reproduce the multilocation repeated-measures
example of Milliken and Johnson (2009, Section 28.3). They contain 153 rows,
including 28 missing responses.

```{r combined-example, eval=FALSE}
library(lmmix)

fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Drug:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ Drug)
```

The data-first interface also works with the base R pipe:

```{r pipe-example, eval=FALSE}
fit <- multicentre |>
  lmm(
    formula = Y ~ Drug * Time,
    random = ~ 1 | Center,
    repeated = ~ Time | Center:Drug:Subject,
    structure = "ar1"
  )
```

## Supported covariance structures

The residual structures are:

* `id`: homogeneous independent residuals;
* `cs`: compound symmetry;
* `ar1`: first-order autoregression, including negative correlation;
* `toep`: positive-definite Toeplitz covariance parameterized through partial
  autocorrelations;
* `un`: unstructured covariance parameterized through a Cholesky factor.

Random intercepts and random slopes use an unstructured Cholesky-parameterized
`G` matrix.

## Validation status

Executable tests compare the package with `nlme` for ML, REML, random
intercepts, random slopes, compound symmetry, and AR(1). A marginal AR(1)
model is also compared with `mmrm`. Satterthwaite tests are compared with
`lmerTest` where the models overlap.

The multilocation validation tests compare fitted covariance parameters,
fixed effects, type III tests, LS-means, pairwise differences, and
Satterthwaite degrees of freedom with stored SAS PROC MIXED benchmarks for the
same model specification. The [validation note](VALIDATION.md) documents the
theoretical sources, model specification, and numerical tolerances.

## References

* Patterson, H. D., and Thompson, R. (1971). Recovery of inter-block
  information when block sizes are unequal. *Biometrika*, 58(3), 545-554.
  <https://doi.org/10.1093/biomet/58.3.545>
* Harville, D. A. (1977). Maximum likelihood approaches to variance component
  estimation and to related problems. *Journal of the American Statistical
  Association*, 72(358), 320-338.
  <https://doi.org/10.1080/01621459.1977.10480998>
* LaMotte, L. R. (2007). A direct derivation of the REML likelihood function.
  *Statistical Papers*, 48(2), 321-327.
  <https://doi.org/10.1007/s00362-006-0335-6>
* Satterthwaite, F. E. (1946). An approximate distribution of estimates of
  variance components. *Biometrics Bulletin*, 2(6), 110-114.
  <https://doi.org/10.2307/3002019>
* Fai, A. H.-T., and Cornelius, P. L. (1996). Approximate F-tests of multiple
  degree of freedom hypotheses in generalized least squares analyses of
  unbalanced split-plot experiments. *Journal of Statistical Computation and
  Simulation*, 54(4), 363-378.
  <https://doi.org/10.1080/00949659608811740>
* Searle, S. R., Speed, F. M., and Milliken, G. A. (1980). Population marginal
  means in the linear model: An alternative to least squares means.
  *The American Statistician*, 34(4), 216-221.
  <https://doi.org/10.1080/00031305.1980.10483031>
* Pinheiro, J. C., and Bates, D. M. (2000). *Mixed-Effects Models in S and
  S-PLUS*. Springer. <https://doi.org/10.1007/b98882>
* Milliken, G. A., and Johnson, D. E. (2009). *Analysis of Messy Data, Volume
  1: Designed Experiments* (2nd ed.), Section 28.3. Chapman and Hall/CRC.
  <https://doi.org/10.1201/EBK1584883340>
