---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# lmmix

`lmmix` fits Gaussian linear mixed models by explicitly optimizing the ML or
REML criterion for

\[
V = ZGZ^\mathsf{T} + R.
\]

The package is designed for models that combine random effects with correlated
residuals. It provides Satterthwaite denominator degrees of freedom, type III
tests, estimated marginal means, pairwise contrasts, broom methods, and
`emmeans` interoperability.

The package name is `lmmix`; the main fitting function is `lmm()`.

## Installation

Install the package from a local checkout:

```{r installation, eval=FALSE}
devtools::install("path/to/lmmix")
```

## Combined random-effect and residual-correlation model

The included `multicentre` data reproduce Table 5.20 of the source thesis.
They contain the 153 rows of the multi-site design, including 28 missing
responses.

```{r combined-example, eval=FALSE}
library(lmmix)

fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Drug:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ Drug)
```

The data-first interface also works with the base R pipe:

```{r pipe-example, eval=FALSE}
fit <- multicentre |>
  lmm(
    formula = Y ~ Drug * Time,
    random = ~ 1 | Center,
    repeated = ~ Time | Center:Drug:Subject,
    structure = "ar1"
  )
```

## Supported covariance structures

The residual structures are:

* `id`: homogeneous independent residuals;
* `cs`: compound symmetry;
* `ar1`: first-order autoregression, including negative correlation;
* `toep`: positive-definite Toeplitz covariance parameterized through partial
  autocorrelations;
* `un`: unstructured covariance parameterized through a Cholesky factor.

Random intercepts and random slopes use an unstructured Cholesky-parameterized
`G` matrix.

## Validation status

Executable tests compare the package with `nlme` for ML, REML, random
intercepts, random slopes, compound symmetry, and AR(1). A marginal AR(1)
model is also compared with `mmrm`. Satterthwaite tests are compared with
`lmerTest` where the models overlap.

The multi-site validation tests compare fitted covariance parameters, fixed
effects, type III tests, LS-means, pairwise differences, and Satterthwaite
degrees of freedom with the reference tables in the source thesis.
The validation vignette documents the data provenance, the published
precision of the reference values, and an erroneous Drug 3 prototype contrast
carried into Annex B.

## References

* Satterthwaite, F. E. (1946). An approximate distribution of estimates of
  variance components. *Biometrics Bulletin*, 2(6), 110-114.
  <https://doi.org/10.2307/3002019>
* Pinheiro, J. C., and Bates, D. M. (2000). *Mixed-Effects Models in S and
  S-PLUS*. Springer. <https://doi.org/10.1007/b98882>
