---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  output_hook(gsub("[ \\t]+(?=\\n|$)", "", x, perl = TRUE), options)
})
```

# lmmix

`lmmix` fits Gaussian linear mixed models by explicitly optimizing the ML or
REML criterion. Its main use case is a model that contains both random effects
and a structured residual covariance:

\[
Y = X\beta + Zu + \varepsilon,
\qquad
V = ZGZ^\mathsf{T} + R.
\]

The package estimates the covariance parameters, fixed effects and empirical
BLUPs in one fitted object. It also provides Satterthwaite inference, type III
tests, estimated marginal means, pairwise contrasts, standard model methods,
`broom` methods and `emmeans` interoperability.

The package is under active development. The table below distinguishes current
functionality from features that are not implemented.

## Installation

The repository is private. Clone it with the configured GitHub SSH key, then
install the local checkout:

```sh
git clone git@github.com:AurelienNicosiaULaval/lmmix.git
R CMD INSTALL lmmix
```

From an existing checkout, installation from R is also possible:

```{r installation-r, eval=FALSE}
devtools::install_local("/path/to/lmmix")
```

## Current capability matrix

| Component | Current behavior |
|:--|:--|
| Response | Univariate continuous Gaussian response |
| Estimation | Explicit profiled ML and REML optimization |
| Random effects | No random effect, random intercept, or correlated random slopes for one grouping formula |
| Residual covariance | `id`, `cs`, `ar1`, `toep`, and `un` |
| Combined covariance | Random effects and any supported residual structure in the same model |
| Denominator degrees of freedom | Satterthwaite or residual |
| Fixed-effect tests | Coefficient tests and type III tests |
| Marginal means | Equal-weight estimated marginal means and automatic pairwise contrasts |
| Multiplicity | Any method accepted by `stats::p.adjust()` for pairwise p-values |
| Interoperability | `emmeans`, `tidy()`, `glance()`, and `augment()` |
| Not implemented | Kenward-Roger inference, likelihood-ratio model comparison, multiple random-effect terms, generalized responses |

## Quick start: random center effect and AR(1) residuals

The included `multicentre` data are a three-center repeated-measures example
with three treatments and three measurement occasions. The data contain 153
rows, including 28 missing responses. Their source is Milliken and Johnson
(2009, Section 28.3).

```{r combined-fit}
library(lmmix)

fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Drug:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

fit
```

`lmm()` uses complete cases across every variable required by the fixed,
random and repeated formulas. The fitted object therefore records 125
observations for this example.

```{r combined-inference}
nobs(fit)
anova(fit)
VarCorr(fit)
```

Estimated marginal means and their pairwise differences are generated from
the fixed-effects reference grid. No contrast rows need to be written by hand.

```{r combined-lsmeans}
lsmeans(fit, ~Drug)
lsmeans(fit, pairwise ~ Drug, adjust = "holm")
```

The convergence code and Hessian diagnostic should be checked before
interpreting inference.

```{r convergence}
fit$convergence[c(
  "code",
  "optimizer",
  "hessian_positive_definite"
)]
```

## Three supported model families

The same function covers three model families.

```{r model-families, eval=FALSE}
# Marginal repeated-measures model
marginal_fit <- lmm(
  data,
  response ~ treatment * time,
  repeated = ~ time | subject,
  structure = "un"
)

# Random-effects model with independent residuals
random_fit <- lmm(
  data,
  response ~ treatment + time,
  random = ~ 1 + time | subject,
  structure = "id"
)

# Combined random-effect and correlated-residual model
combined_fit <- lmm(
  data,
  response ~ treatment * time,
  random = ~ 1 | center,
  repeated = ~ time | center:subject,
  structure = "ar1"
)
```

For `cs`, `ar1`, `toep`, or `un`, `repeated` must identify one ordering
variable and one grouping expression. Each repeated-measures group may contain
at most one observation at a given ordering value.

## Working with a fitted model

Standard methods return fixed effects, random effects, covariance components,
predictions, residuals and information criteria.

```{r model-methods}
fixef(fit)
head(ranef(fit))
head(residuals(fit))
head(predict(fit))
```

The `generics` methods return tibbles suitable for downstream analysis.

```{r broom-methods}
generics::tidy(fit)
generics::glance(fit)
head(generics::augment(fit))
```

When `emmeans` is installed, the package supplies the fixed-effects basis and
model-specific denominator degrees of freedom.

```{r emmeans-example, eval=requireNamespace("emmeans", quietly = TRUE)}
emmeans::emmeans(fit, ~Drug)
```

## Validation

The test suite separates numerical agreement from general software tests.

| Model component | External comparison | Quantities checked |
|:--|:--|:--|
| Random intercept and random slope | `nlme::lme()` | Fixed effects, covariance parameters, log-likelihood |
| Marginal CS and AR(1) | `nlme::gls()` | Fixed effects, covariance parameters, log-likelihood |
| Marginal AR(1) | `mmrm::mmrm()` | Fixed effects, residual covariance, log-likelihood |
| Satterthwaite inference | `lmerTest::lmer()` | Coefficient degrees of freedom and type III tests |
| Combined random center and AR(1) | Stored PROC MIXED targets | Covariance parameters, fixed effects, type III tests and marginal means |

The package also tests positive definiteness and convergence for every
supported residual structure in a combined model. See `VALIDATION.md` and the
validation vignette for model specifications, tolerances and limitations.

## Current limitations

The following limitations are part of the current `0.1.0` interface:

* `ddf = "kenward-roger"` is rejected because Kenward-Roger inference is not
  implemented.
* `anova()` provides type III tests for one fitted model. It does not compare
  nested fitted models.
* Only one random-effects formula is accepted. Multiple crossed random-effect
  terms are not implemented.
* The implementation constructs a dense marginal covariance matrix. It is
  intended for small and moderate data sets, not large-scale sparse problems.
* Missing observations are removed by complete-case analysis over all model
  variables.
* Pairwise multiplicity adjustments modify p-values. The reported confidence
  intervals remain pointwise intervals.
* New random-effect groups receive a random contribution of zero in
  conditional prediction.

## Theoretical sources

The REML implementation follows Patterson and Thompson (1971), Harville
(1977), and LaMotte (2007). Satterthwaite inference follows Satterthwaite
(1946), with multi-degree-of-freedom tests based on Fai and Cornelius (1996).
Estimated marginal means follow Searle, Speed, and Milliken (1980). Full
citations and DOI links are included in the package help and vignettes.
