---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  output_hook(gsub("[ \\t]+(?=\\n|$)", "", x, perl = TRUE), options)
})
```

# lmmix

`lmmix` fits Gaussian linear mixed models by explicitly optimizing the ML or
REML criterion. Its main use case is a model that contains both random effects
and a structured residual covariance.

The package estimates the covariance parameters, fixed effects and empirical
BLUPs in one fitted object. It also provides Satterthwaite and Kenward-Roger
inference, type III tests, likelihood-ratio comparisons, estimated marginal
means, pairwise contrasts, standard model methods, `broom` methods and
`emmeans` interoperability.

The package is under active development. The table below summarizes the
implemented functionality and the intended model boundary.

## Installation

Install the development version directly from the public GitHub repository:

```{r installation, eval=FALSE}
install.packages("devtools")
devtools::install_github("AurelienNicosiaULaval/lmmix")
```

## Vignettes

The package contains three complementary vignettes:

* [Using lmmix](articles/lmmix.html)
  ([source](vignettes/lmmix.Rmd))
  presents the fitting workflow and the main methods;
* [Estimation and inference in lmmix](articles/theory-reml.html)
  ([source](vignettes/theory-reml.Rmd))
  gives the mathematical formulation, covariance parameterizations, ML and
  REML criteria, and Satterthwaite and Kenward-Roger inference;
* [Validation against independent R implementations and PROC MIXED](articles/validation.html)
  ([source](vignettes/validation.Rmd))
  documents the independent comparisons, numerical tolerances, and validation
  boundaries.

They are also available from an installed package:

```{r open-vignettes, eval=FALSE}
vignette("lmmix", package = "lmmix")
vignette("theory-reml", package = "lmmix")
vignette("validation", package = "lmmix")
```

## Current capability matrix

| Component | Current behavior |
|:--|:--|
| Response | Univariate continuous Gaussian response |
| Estimation | Explicit profiled ML and REML optimization |
| Random effects | No random effect, one grouping formula, or independent crossed or nested formulas with correlated slopes within each term |
| Residual covariance | `id`, `cs`, `ar1`, `toep`, and `un` |
| Combined covariance | Random effects and any supported residual structure in the same model |
| Denominator degrees of freedom | Satterthwaite, Kenward-Roger, or residual |
| Fixed-effect tests | Coefficient tests, type III tests, and nested-model likelihood-ratio comparisons |
| Marginal means | Equal-weight estimated marginal means and automatic pairwise contrasts |
| Multiplicity | Any `stats::p.adjust()` method and simultaneous Bonferroni confidence intervals |
| Missing data | `na.omit`, `na.exclude`, or `na.fail` |
| Interoperability | `emmeans`, `tidy()`, `glance()`, and `augment()` |
| Diagnostic plots | `plot.lmm()` returns `ggplot2` residual, Q-Q, and fitted-value diagnostics |
| Model boundary | Gaussian responses and dense covariance factorization for small and moderate data sets |

## Quick start: random center effect and AR(1) residuals

The included `multicentre` data are a three-center repeated-measures example
with three treatments and three measurement occasions. The data contain 153
rows, including 28 missing responses. Their source is Milliken and Johnson
(2009, Section 28.3).

```{r combined-fit}
library(lmmix)

fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Drug:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

fit
```

By default, `lmm()` omits rows that are incomplete for variables required by
the fixed, random and repeated formulas. The fitted object therefore records
125 observations for this example. Use `na.action = na.exclude` to restore
omitted positions in fitted values, residuals and augmented data.

```{r combined-inference}
nobs(fit)
anova(fit)
VarCorr(fit)
```

Estimated marginal means and their pairwise differences are generated from
the fixed-effects reference grid. No contrast rows need to be written by hand.

```{r combined-lsmeans}
lsmeans(fit, ~Drug)
lsmeans(fit, pairwise ~ Drug, adjust = "holm")
```

The convergence code and Hessian diagnostic should be checked before
interpreting inference.

```{r convergence}
fit$convergence[c(
  "code",
  "optimizer",
  "hessian_positive_definite"
)]
```

## Three supported model families

The same function covers three model families.

```{r model-families, eval=FALSE}
# Marginal repeated-measures model
marginal_fit <- lmm(
  data,
  response ~ treatment * time,
  repeated = ~ time | subject,
  structure = "un"
)

# Random-effects model with independent residuals
random_fit <- lmm(
  data,
  response ~ treatment + time,
  random = ~ 1 + time | subject,
  structure = "id"
)

# Crossed random intercepts
crossed_fit <- lmm(
  data,
  response ~ treatment + time,
  random = list(
    center = ~ 1 | center,
    subject = ~ 1 | subject
  )
)

# Combined random-effect and correlated-residual model
combined_fit <- lmm(
  data,
  response ~ treatment * time,
  random = ~ 1 | center,
  repeated = ~ time | center:subject,
  structure = "ar1"
)
```

For `cs`, `ar1`, `toep`, or `un`, `repeated` must identify one ordering
variable and one grouping expression. Each repeated-measures group may contain
at most one observation at a given ordering value.

## Working with a fitted model

Standard methods return fixed effects, random effects, covariance components,
predictions, residuals and information criteria.

```{r model-methods}
fixef(fit)
head(ranef(fit))
head(residuals(fit))
head(predict(fit))
```

The `generics` methods return tibbles suitable for downstream analysis.
Printed tables use readable headings without dots, including `Std Error`,
`p value`, `Conf Low`, and `Conf High`. The underlying objects retain the
standard `broom` names required for programmatic use.

```{r broom-methods}
generics::tidy(fit)
generics::glance(fit)
head(generics::augment(fit))
```

The S3 `plot()` method returns a `ggplot2` object and supports residual,
normal Q-Q, and observed-versus-fitted diagnostics.

```{r diagnostic-plot, fig.alt="Standardized residuals plotted against fitted values"}
plot(fit, which = "residuals")
```

When `emmeans` is installed, the package supplies the fixed-effects basis and
model-specific denominator degrees of freedom.

```{r emmeans-example, eval=requireNamespace("emmeans", quietly = TRUE)}
emmeans::emmeans(fit, ~Drug)
```

Kenward-Roger inference is available for REML fits. Nested models with
different fixed effects are automatically refitted with ML for a
likelihood-ratio comparison.

```{r extended-inference, eval=FALSE}
kr_fit <- lmm(
  data,
  response ~ treatment * time,
  random = ~ 1 | subject,
  ddf = "kenward-roger"
)

anova(reduced_fit, full_fit)
```

## Validation

The test suite separates numerical agreement from general software tests.

| Model component | External comparison | Quantities checked |
|:--|:--|:--|
| Random intercept and random slope | `nlme::lme()` | Fixed effects, covariance parameters, log-likelihood |
| Marginal CS and AR(1) | `nlme::gls()` | Fixed effects, covariance parameters, log-likelihood |
| Marginal AR(1) | `mmrm::mmrm()` | Fixed effects, residual covariance, log-likelihood |
| Satterthwaite inference | `lmerTest::lmer()` | Coefficient degrees of freedom and type III tests |
| Kenward-Roger inference | `lmerTest` and `mmrm` | Adjusted covariance and denominator degrees of freedom |
| Crossed random effects | `lme4::lmer()` | Fixed effects, covariance parameters and log-likelihood |
| Nested-model comparisons | `lme4::anova()` | Likelihood-ratio statistics and p-values |
| Official split-plot example 79.1 | Published PROC MIXED output | Variance components, REML criterion and type III tests |
| Official repeated-measures example 79.2 | Published PROC MIXED output | UN and CS covariance, ML criterion, fixed effects and type III statistics |
| Official random-coefficients example 79.5 | Published PROC MIXED output | Random covariance, REML criterion, fixed effects, BLUPs and type III test |
| Combined random center and AR(1) | Stored PROC MIXED targets | Covariance parameters, fixed effects, type III tests and marginal means |

The package also tests positive definiteness and convergence for every
supported residual structure in a combined model. See `VALIDATION.md` and the
validation vignette for model specifications, tolerances and limitations.

## Model scope and remaining boundaries

Version `0.2.1` fits univariate Gaussian models. It is not a generalized
mixed-model engine. The likelihood constructs and factors a dense marginal
covariance matrix, so the package targets small and moderate data sets rather
than large-scale sparse problems.

Rows with missing responses or model covariates are handled according to
`na.action`; the package does not impute missing covariates. Conditional
prediction rejects unseen random-effect groups by default. Set
`allow.new.levels = TRUE` to request the population-level prediction, for
which the unseen random contribution is zero.

## Theoretical sources

The REML implementation follows Patterson and Thompson (1971), Harville
(1977), and LaMotte (2007). Satterthwaite inference follows Satterthwaite
(1946), with multi-degree-of-freedom tests based on Fai and Cornelius (1996).
Small-sample covariance adjustment follows Kenward and Roger (1997).
Estimated marginal means follow Searle, Speed, and Milliken (1980). Full
citations and DOI links are included in the package help and the
[theory vignette](articles/theory-reml.html).

## Package citation

The installed package provides its current academic citation directly:

```r
citation("lmmix")
```
