---
title: "Covariance intervals and boundary-aware model comparisons"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Covariance intervals and boundary-aware model comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(lmmix)
library(ggplot2)
```

## Scope

This vignette covers two inferential questions that differ from inference on
fixed effects:

1. confidence intervals for variance and correlation parameters;
2. likelihood-ratio comparisons when the null model can place a variance on
   the boundary at zero.

The usual chi-squared likelihood-ratio reference can fail at a boundary
[@self1987]. A parametric bootstrap instead estimates the null distribution by
simulating from the fitted smaller model and refitting both models to every
simulated response [@davison1997].

## Confidence intervals for covariance parameters

The `theta_` selector requests all covariance parameters. Variance intervals
are formed on the log scale, and correlation intervals are formed on the
Fisher-z scale. The delta-method standard errors are transformed before the
normal critical value is applied. This keeps variance limits positive and
correlation limits inside `(-1, 1)`.

```{r covariance-intervals}
orthodont <- as.data.frame(nlme::Orthodont)

fit <- lmm(
  orthodont,
  distance ~ age + Sex,
  random = ~ age | Subject
)

confint(fit, parm = "theta_")
```

The default `confint(fit)` remains backward compatible and returns only fixed
effects. Use `parm = c("beta_", "theta_")` to request both families. These are
local Wald intervals based on the observed likelihood Hessian. They should not
be interpreted as profile-likelihood or bootstrap intervals, especially when
the Hessian is poorly conditioned or an estimate is close to a boundary.

## Parametric-bootstrap likelihood-ratio test

The following comparison asks whether the Orthodont data support a random
subject intercept beyond an independent Gaussian residual model. Both models
are fitted by ML because covariance-model likelihood-ratio comparisons must
use the same fixed effects under ML.

```{r boundary-models}
null_fit <- lmm(
  orthodont,
  distance ~ age + Sex,
  method = "ML"
)

random_fit <- lmm(
  orthodont,
  distance ~ age + Sex,
  random = ~ 1 | Subject,
  method = "ML"
)
```

The ordinary call retains the asymptotic chi-squared reference and emits a
boundary warning. The bootstrap alternative is requested explicitly.

```{r boundary-bootstrap}
asymptotic <- suppressWarnings(anova(null_fit, random_fit))
bootstrap <- anova(
  null_fit,
  random_fit,
  test = "parametric.bootstrap",
  nsim = 49,
  seed = 20260714
)

data.frame(
  method = c("asymptotic chi-squared", "parametric bootstrap"),
  statistic = c(asymptotic$Chisq[2], bootstrap$Chisq[2]),
  p_value = c(asymptotic$p.value[2], bootstrap$p.value[2])
)
```

For each simulation, `lmmix` draws a Gaussian response using the fitted mean
and marginal covariance of the null model, refits both models by ML, and
recomputes the likelihood-ratio statistic. The finite-simulation correction is

\[
\widehat p = \frac{1 + \sum_{b=1}^{B}
I(T_b \geq T_{\mathrm{obs}})}{B + 1}.
\]

The simulated null distribution is retained as an attribute for diagnostics.

```{r boundary-distribution}
bootstrap_statistics <- attr(bootstrap, "bootstrap")[[2]]$statistics

ggplot(
  data.frame(statistic = bootstrap_statistics),
  aes(x = statistic)
) +
  geom_histogram(bins = 12, color = "white", fill = "#2C7FB8") +
  geom_vline(
    xintercept = bootstrap$Chisq[2],
    color = "#D7301F",
    linewidth = 0.9
  ) +
  labs(
    x = "Simulated likelihood-ratio statistic under the null",
    y = "Count",
    title = "Parametric-bootstrap reference distribution"
  ) +
  theme_minimal(base_size = 12)
```

The 49 simulations above keep the installed vignette reasonably fast and give
a p-value resolution of `1 / 50`. They are a reproducible demonstration, not a
publication-grade Monte Carlo analysis. For substantive inference, increase
`nsim`, commonly to at least 999, and examine the retained distribution and
the number of successful refits. The returned p-value uses only successful
refits and the method warns if any refit fails.

## Reproducibility

Supplying `seed` makes the simulated reference distribution reproducible and
restores the caller's random-number state on exit. Omitting `seed` uses and
advances the current R random-number stream. The fitted objects, simulation
count, successful-refit count, and simulated statistics remain available in
the returned comparison object.

```{r session-info}
sessionInfo()
```
