---
title: "Getting started with lmmix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with lmmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`lmmix` fits the Gaussian model

\[
Y = X\beta + Zu + \varepsilon,
\qquad
V = ZGZ^\mathsf{T} + R,
\]

by directly optimizing a profiled ML or REML objective.

```{r}
library(lmmix)
```

## A model with two covariance sources

The example data contain a random center effect and repeated observations
within subjects.

```{r}
fit <- lmm(
  multicentre,
  Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Subject,
  structure = "ar1"
)

fit
```

The summary contains coefficient-level tests, type III tests, covariance
parameters, and information criteria.

```{r}
summary(fit)
```

## Estimated marginal means

Nuisance-factor levels receive equal weights. Numeric covariates not requested
in `specs` are held at their observed means.

```{r}
lsmeans(fit, ~Drug)
lsmeans(fit, pairwise ~ Drug)
```

## Tidy output

```{r}
generics::tidy(fit)
generics::glance(fit)
head(generics::augment(fit))
```

When `emmeans` is installed, the package supplies the required basis and uses
the model's Satterthwaite degrees of freedom.

```{r, eval=requireNamespace("emmeans", quietly = TRUE)}
emmeans::emmeans(fit, ~Drug)
```
