---
title: "Using lmmix"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Using lmmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Model scope

`lmmix` fits Gaussian linear mixed models with random effects, structured
residual covariance, or both components in the same model. The package
directly optimizes a profiled ML or REML objective and does not delegate
estimation to another mixed-model package.

This vignette focuses on practical use. The mathematical formulation,
covariance parameterizations, likelihood criteria, and inference derivations
are presented separately in the vignette *Estimation and inference in lmmix*.

```{r}
library(lmmix)
```

## Prepare the data

The included `multicentre` data have three treatments, three measurement
occasions, three centers and subjects nested within center and treatment.

```{r data-summary}
str(multicentre)
colSums(is.na(multicentre))
```

The fixed, random and repeated formulas jointly determine the variables used
by the analysis. `lmm()` removes rows that are incomplete for any of those
variables and stores the retained data in the fitted object.

## Fit a combined covariance model

This model contains a random center intercept and an AR(1) residual covariance
within subject.

```{r combined-fit}
fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Drug:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

fit
```

The three formulas have distinct roles:

| Argument | Role in the fitted model |
|:--|:--|
| `formula` | Defines the response and fixed-effects matrix `X` |
| `random` | Defines the random-effects matrix `Z` and grouping factor for `G` |
| `repeated` | Defines the ordering variable and independent blocks for `R` |

For the residual structures other than `id`, each repeated-measures group must
have at most one observation per ordering value.

## Inspect convergence before inference

The optimizer code and Hessian diagnostic are part of the fitted object.

```{r convergence}
fit$convergence[c(
  "code",
  "message",
  "optimizer",
  "iterations",
  "hessian_positive_definite"
)]
```

A nonzero convergence code or a non-positive-definite likelihood Hessian
requires investigation before interpreting standard errors or denominator
degrees of freedom. The fitting function emits a warning in either case.

## Fixed effects and type III tests

`summary()` combines coefficient-level inference, type III tests, covariance
parameters and information criteria.

```{r summary}
summary(fit)
```

Individual components are also available through standard methods.

```{r inference}
fixef(fit)
vcov(fit)
anova(fit)
confint(fit)
```

`anova()` currently performs type III fixed-effect tests for one model. It does
not compare nested models. Satterthwaite and residual denominator degrees of
freedom are implemented. Kenward-Roger inference is not implemented.

## Covariance parameters and random effects

`VarCorr()` returns the estimated covariance components on their natural
scale. `ranef()` returns empirical BLUPs as a tibble.

```{r covariance}
VarCorr(fit)
ranef(fit)
```

For a random-slope model, the random-effect covariance is unstructured. Its
variances and correlations appear as separate rows in `VarCorr()`.

## Estimated marginal means

Marginal means use equal weights over nuisance-factor levels. Numeric
covariates not requested in `specs` are held at their observed means. This
construction targets population marginal means in the sense of @searle1980.

```{r lsmeans}
lsmeans(fit, ~Drug)
lsmeans(fit, ~Time)
```

Pairwise contrast matrices are generated automatically. The `adjust` argument
is passed to `stats::p.adjust()` and therefore adjusts pairwise p-values. The
confidence intervals remain pointwise.

```{r pairwise}
lsmeans(fit, pairwise ~ Drug, adjust = "holm")
```

Reference values for numeric covariates can be overridden with `at`:

```{r at-example, eval=FALSE}
lsmeans(fit, ~Drug, at = list(Time = "2"))
```

When `emmeans` is installed, `lmmix` supplies the required fixed-effects basis
and model-specific denominator degrees of freedom.

```{r emmeans, eval=requireNamespace("emmeans", quietly = TRUE)}
emmeans::emmeans(fit, ~Drug)
```

## Predictions, residuals and tidy output

Conditional fitted values include empirical random effects. Marginal fitted
values use fixed effects only.

```{r predictions}
head(fitted(fit, type = "conditional"))
head(fitted(fit, type = "marginal"))
head(residuals(fit, type = "pearson"))
```

For `predict(fit, newdata)`, known grouping levels receive their fitted random
effect. New groups receive a random contribution of zero. Set `re.form` to a
non-`NULL` value to request fixed-effects-only predictions.

The `generics` methods return tibbles.

```{r broom}
generics::tidy(fit)
generics::glance(fit)
head(generics::augment(fit))
```

## Alternative model specifications

The same interface covers marginal, random-effects and combined models.

```{r alternatives, eval=FALSE}
# Marginal model with unstructured repeated-measures covariance
fit_marginal <- lmm(
  data,
  response ~ treatment * time,
  repeated = ~ time | subject,
  structure = "un"
)

# Random intercept and slope with independent residuals
fit_random <- lmm(
  data,
  response ~ treatment + time,
  random = ~ 1 + time | subject,
  structure = "id"
)

# Combined random intercept and Toeplitz residual covariance
fit_combined <- lmm(
  data,
  response ~ treatment * time,
  random = ~ 1 | center,
  repeated = ~ time | center:subject,
  structure = "toep"
)
```

## Practical limits

The current implementation constructs the full dense matrix `V` during
optimization. It is appropriate for small and moderate data sets. It is not a
large-scale sparse mixed-model engine. The interface accepts one random-effects
formula and one repeated-measures structure. Generalized responses, multiple
random-effect terms, Kenward-Roger inference and likelihood-ratio model
comparison are outside the current `0.1.0` implementation.
