---
title: "Multi-site PROC MIXED validation workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-site PROC MIXED validation workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Scope of the available data

Annex B describes a three-center, three-drug, three-time design and provides
selected PROC MIXED outputs. It does not provide the original observation
table. The package therefore includes `multicentre`, a simulated data set with
the same design characteristics and several missing responses.

The simulated data cannot validate the numerical SAS targets. Tests for those
targets are present in `tests/testthat/test-validation-sas.R` and are skipped
with this reason. They can only be activated after the original observations
are supplied.

## End-to-end model

```{r}
library(lmmix)

fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

fit
VarCorr(fit)
anova(fit)
```

Treatment and time LS-means and treatment differences are generated without
hand-written contrast rows.

```{r}
lsmeans(fit, ~Drug)
lsmeans(fit, ~Time)
lsmeans(fit, pairwise ~ Drug)
```

## Targets awaiting original data

The following values are transcribed from Annex B and are not results from the
simulated data.

| Quantity | Level or contrast | Estimate | Standard error | df |
|:--|:--|--:|--:|--:|
| Drug LS-mean | 1 | 13.156900 | 1.529980 | 2.97 |
| Drug LS-mean | 2 | 18.148460 | 1.533334 | 3.00 |
| Drug LS-mean | 3 | 16.254250 | 1.540690 | 3.06 |
| Time LS-mean | 1 | 14.954420 | 1.397169 | 2.07 |
| Time LS-mean | 2 | 15.670090 | 1.399488 | 2.09 |
| Time LS-mean | 3 | 17.732510 | 1.404265 | 2.12 |
| Drug difference | 1 minus 2 | -4.991562 | 1.099005 | 46.10 |

The Annex B two-sided p-value for the last contrast is `4.0e-05`. No tolerance
claim is made until the original data are available.
