---
title: "Multi-site PROC MIXED validation workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-site PROC MIXED validation workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Scope of the available data

Annex B describes a three-center, three-drug, three-time design and provides
selected PROC MIXED outputs. The complete observation table was recovered
from Table 5.20 of the locally supplied final thesis manuscript. The included
`multicentre` data reproduce its 153 rows and 28 missing responses. The fitted
model consequently uses 125 complete responses, as reported in Table 5.6.

## End-to-end model

```{r}
library(lmmix)

fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Drug:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

fit
VarCorr(fit)
anova(fit)
```

Treatment and time LS-means and treatment differences are generated without
hand-written contrast rows.

```{r}
lsmeans(fit, ~Drug)
lsmeans(fit, ~Time)
lsmeans(fit, pairwise ~ Drug)
```

## Reference targets

The following values follow the SAS column of Table 5.17, which aligns the
levels with the ascending coding in Table 5.20. Table 5.11 contains the same
sets of drug and time estimates, but displays the outer levels in reverse
order. The tests use the data-aligned level labels.

| Quantity | Level or contrast | Estimate | Standard error | df |
|:--|:--|--:|--:|--:|
| Drug LS-mean | 1 | 13.1568 | 1.5290 | 2.98 |
| Drug LS-mean | 2 | 18.1484 | 1.5324 | 3.01 |
| Drug LS-mean | 3 | 17.0516 | 1.5327 | 3.01 |
| Time LS-mean | 1 | 14.9543 | 1.3961 | 2.08 |
| Time LS-mean | 2 | 15.6700 | 1.3985 | 2.09 |
| Time LS-mean | 3 | 17.7324 | 1.4032 | 2.12 |
| Drug difference | 1 minus 2 | -4.9916 | 1.0990 | 46.1 |

The two-sided p-value for the last contrast is below `0.0001`; the package
computes `3.998e-05` before rounding.

Table 5.18 supplies the remaining pairwise references. With contrasts oriented
in ascending level order, they are:

| Factor | Contrast | Estimate | Standard error | df | p-value |
|:--|:--|--:|--:|--:|--:|
| Drug | 1 minus 3 | -3.8948 | 1.0987 | 46.1 | 0.0009 |
| Drug | 2 minus 3 | 1.0968 | 1.1043 | 46.9 | 0.3257 |
| Time | 1 minus 2 | -0.7157 | 0.1845 | 68.1 | 0.0002 |
| Time | 1 minus 3 | -2.7781 | 0.2714 | 73.2 | below 0.0001 |
| Time | 2 minus 3 | -2.0624 | 0.2056 | 68.6 | below 0.0001 |

## Source erratum and numerical precision

Annex B reports `16.25425` for the Drug 3 LS-mean. This is the result of the
prototype contrast on p. 109, not the SAS LS-mean. That contrast omits one
third of the `Time3` main-effect coefficient. In the fitted coefficient order,
the prototype row is

```{r, eval=FALSE}
c(1, 0, 1, 1 / 3, 0, 0, 1 / 3, 0, 1 / 3)
```

The automatically constructed Drug 3 LS-mean includes all three time levels
and gives `17.05158`, agreeing with the SAS value `17.0516`. An executable
regression test preserves the traceability of the prototype value without
returning it as a marginal mean.

Some source degrees of freedom are printed with only one or two decimal
places. Such printed values do not contain enough precision for an absolute
`1e-3` comparison. Tests therefore compare estimates and standard errors at
absolute tolerance `1e-3`, and compare degrees of freedom at their published
precision.
