---
title: "Multi-site PROC MIXED validation workflow"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Multi-site PROC MIXED validation workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Scope of the available data

The `multicentre` data reproduce the multilocation repeated-measures example
described by @milliken2009 in Section 28.3. The design has three centers,
three drugs, three measurement times, and 153 rows. There are 28 missing
responses, so the fitted model uses 125 complete responses.

## End-to-end model

```{r}
library(lmmix)

fit <- lmm(
  data = multicentre,
  formula = Y ~ Drug * Time,
  random = ~ 1 | Center,
  repeated = ~ Time | Center:Drug:Subject,
  structure = "ar1",
  method = "REML",
  ddf = "satterthwaite"
)

fit
VarCorr(fit)
anova(fit)
```

Treatment and time LS-means and treatment differences are generated without
hand-written contrast rows.

```{r}
lsmeans(fit, ~Drug)
lsmeans(fit, ~Time)
lsmeans(fit, pairwise ~ Drug)
```

## PROC MIXED benchmark

The stored reference values correspond to the following PROC MIXED
specification. The RANDOM and REPEATED statements represent the two covariance
sources, and
`DDFM=SATTERTHWAITE` requests row-specific denominator degrees of freedom as
documented by @sas2026.

```sas
proc mixed data=multicentre method=reml;
  class Center Drug Subject Time;
  model Y = Drug Time Drug*Time / ddfm=satterth solution;
  random Center;
  repeated Time / type=ar(1) subject=Subject(Center*Drug);
  lsmeans Drug Time / diff;
run;
```

The stored SAS reference values, with levels oriented in ascending data order,
are:

| Quantity | Level or contrast | Estimate | Standard error | df |
|:--|:--|--:|--:|--:|
| Drug LS-mean | 1 | 13.1568 | 1.5290 | 2.98 |
| Drug LS-mean | 2 | 18.1484 | 1.5324 | 3.01 |
| Drug LS-mean | 3 | 17.0516 | 1.5327 | 3.01 |
| Time LS-mean | 1 | 14.9543 | 1.3961 | 2.08 |
| Time LS-mean | 2 | 15.6700 | 1.3985 | 2.09 |
| Time LS-mean | 3 | 17.7324 | 1.4032 | 2.12 |
| Drug difference | 1 minus 2 | -4.9916 | 1.0990 | 46.1 |

The two-sided p-value for the last contrast is below `0.0001`; the package
computes `3.998e-05` before rounding.

The remaining pairwise references are:

| Factor | Contrast | Estimate | Standard error | df | p-value |
|:--|:--|--:|--:|--:|--:|
| Drug | 1 minus 3 | -3.8948 | 1.0987 | 46.1 | 0.0009 |
| Drug | 2 minus 3 | 1.0968 | 1.1043 | 46.9 | 0.3257 |
| Time | 1 minus 2 | -0.7157 | 0.1845 | 68.1 | 0.0002 |
| Time | 1 minus 3 | -2.7781 | 0.2714 | 73.2 | below 0.0001 |
| Time | 2 minus 3 | -2.0624 | 0.2056 | 68.6 | below 0.0001 |

## Numerical precision

The benchmark degrees of freedom are retained at the precision shown by PROC
MIXED. Tests compare estimates and standard errors with absolute tolerance
`1e-3`, and degrees of freedom at the displayed precision. The implementation
is also checked against `nlme`, `lmerTest`, and `mmrm` in model classes where
those packages overlap with `lmmix`.
