---
title: "Validation against independent R implementations and PROC MIXED"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Validation against independent R implementations and PROC MIXED}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

has_nlme <- requireNamespace("nlme", quietly = TRUE)
has_lmer_test <- requireNamespace("lmerTest", quietly = TRUE)
has_mmrm <- requireNamespace("mmrm", quietly = TRUE)

comparison_table <- function(term, lmmix, reference) {
  tibble::tibble(
    term = term,
    lmmix = as.numeric(lmmix),
    reference = as.numeric(reference),
    abs.diff = abs(as.numeric(lmmix) - as.numeric(reference))
  )
}

show_table <- function(x, digits = 6) {
  names(x) <- trimws(gsub("\\.+", " ", names(x)))
  knitr::kable(x, digits = digits, row.names = FALSE)
}
```

## Purpose and evidentiary standard

This vignette reports the numerical comparisons used to validate `lmmix`.
Every comparison is restricted to a model class for which the two
implementations use the same fixed-effects specification, covariance model,
and estimation criterion. A small difference is meaningful only after those
definitions have been aligned.

Validation is organized in three layers:

1. comparisons with independent R implementations for overlapping models;
2. structural checks for combined random-effect and residual-correlation
   models that no single reference R function covers;
3. regression comparisons with stored PROC MIXED targets for one fully
   specified combined model.

The numerical tables below are produced by executable R code when the
vignette is built. The PROC MIXED values are stored targets. The repository
does not contain a SAS execution transcript, so the SAS section is not
presented as a new independent SAS run.

## Comparison map

| lmmix model | reference function | quantities compared | tolerance |
|:--|:--|:--|:--|
| Random-intercept REML | `nlme::lme()` | fixed effects and log-likelihood: `1e-6`; variance components: `1e-4` | absolute |
| Random-slope REML | `nlme::lme()` | fixed effects and log-likelihood: `1e-5`; covariance parameters: `1e-4` | absolute |
| Random-intercept ML | `nlme::lme()` | fixed effects, log-likelihood | `1e-6` |
| Marginal CS and AR(1) REML | `nlme::gls()` | fixed effects, residual variance, correlation, log-likelihood | `1e-5` |
| Marginal AR(1) REML | `mmrm::mmrm()` | fixed effects and log-likelihood: `1e-5`; residual covariance matrix: `1e-4` | absolute |
| Random-intercept REML | `lmerTest::lmer()` | estimates and standard errors: `1e-5`; coefficient df: `5e-3` | absolute |
| Type III Satterthwaite tests | `lmerTest::lmer()` | numerator df: `1e-6`; denominator df and F statistics: `5e-3` | absolute |
| Random center plus AR(1) residual | PROC MIXED stored targets | covariance parameters, fixed effects, type III tests, LS-means, differences | target-specific precision |

`nlme` provides independent likelihood comparisons for random-effects and
structured marginal models [@pinheiro2000; @nlme2025]. `lmerTest` provides an
independent Satterthwaite implementation for models fitted by `lme4`
[@kuznetsova2017]. `mmrm` provides a second independent comparison for the
marginal repeated-measures model [@mmrm2026].

## Shared longitudinal data

The R comparisons use the `Orthodont` data distributed with `nlme`. `Occasion`
is created explicitly so that AR(1) visits have a declared order.

```{r orthodont-data, eval=has_nlme}
library(lmmix)

orthodont <- as.data.frame(nlme::Orthodont)
orthodont$Occasion <- factor(ave(
  seq_len(nrow(orthodont)),
  orthodont$Subject,
  FUN = seq_along
))

tibble::tibble(
  n.obs = nrow(orthodont),
  n.subjects = nlevels(orthodont$Subject),
  n.occasions = nlevels(orthodont$Occasion)
) |>
  show_table(digits = 0)
```

## Random-intercept REML comparison with `nlme::lme()`

The first comparison uses the same fixed formula, random intercept, and REML
criterion in both implementations.

```{r random-intercept-reml, eval=has_nlme}
fit.ri <- lmm(
  orthodont,
  distance ~ age + Sex,
  random = ~ 1 | Subject,
  method = "REML"
)

ref.ri <- nlme::lme(
  distance ~ age + Sex,
  random = ~ 1 | Subject,
  data = orthodont,
  method = "REML"
)

comparison_table(
  term = names(fixef(fit.ri)),
  lmmix = fixef(fit.ri),
  reference = nlme::fixef(ref.ri)
) |>
  show_table()
```

```{r random-intercept-reml-covariance, eval=has_nlme}
ref.ri.cov <- as.matrix(nlme::getVarCov(
  ref.ri,
  type = "random.effects"
))
ref.ri.variance <- c(
  ref.ri.cov[1L, 1L],
  ref.ri$sigma^2
)

comparison_table(
  term = c("random.intercept.var", "residual.var"),
  lmmix = VarCorr(fit.ri)$estimate,
  reference = ref.ri.variance
) |>
  show_table()

comparison_table(
  term = "logLik",
  lmmix = logLik(fit.ri),
  reference = logLik(ref.ri)
) |>
  show_table()
```

## Random-slope REML comparison with `nlme::lme()`

This comparison adds a correlated random slope for age. It checks two random
variances, their correlation, the residual variance, the fixed effects, and
the REML log-likelihood.

```{r random-slope-reml, eval=has_nlme}
fit.rs <- lmm(
  orthodont,
  distance ~ age + Sex,
  random = ~ 1 + age | Subject,
  method = "REML"
)

ref.rs <- nlme::lme(
  distance ~ age + Sex,
  random = ~ age | Subject,
  data = orthodont,
  method = "REML"
)

comparison_table(
  term = names(fixef(fit.rs)),
  lmmix = fixef(fit.rs),
  reference = nlme::fixef(ref.rs)
) |>
  show_table()
```

```{r random-slope-reml-covariance, eval=has_nlme}
ref.rs.cov <- as.matrix(nlme::getVarCov(
  ref.rs,
  type = "random.effects"
))
ref.rs.parameters <- c(
  ref.rs.cov[1L, 1L],
  ref.rs.cov[2L, 2L],
  ref.rs.cov[1L, 2L] / sqrt(ref.rs.cov[1L, 1L] * ref.rs.cov[2L, 2L]),
  ref.rs$sigma^2
)

comparison_table(
  term = c(
    "random.intercept.var",
    "random.age.var",
    "random.intercept.age.cor",
    "residual.var"
  ),
  lmmix = VarCorr(fit.rs)$estimate,
  reference = ref.rs.parameters
) |>
  show_table()

comparison_table(
  term = "logLik",
  lmmix = logLik(fit.rs),
  reference = logLik(ref.rs)
) |>
  show_table()
```

## Maximum-likelihood comparison with `nlme::lme()`

The ML comparison verifies that agreement is not specific to the REML
criterion.

```{r random-intercept-ml, eval=has_nlme}
fit.ml <- lmm(
  orthodont,
  distance ~ age + Sex,
  random = ~ 1 | Subject,
  method = "ML"
)

ref.ml <- nlme::lme(
  distance ~ age + Sex,
  random = ~ 1 | Subject,
  data = orthodont,
  method = "ML"
)

rbind(
  comparison_table(
    term = names(fixef(fit.ml)),
    lmmix = fixef(fit.ml),
    reference = nlme::fixef(ref.ml)
  ),
  comparison_table(
    term = "logLik",
    lmmix = logLik(fit.ml),
    reference = logLik(ref.ml)
  )
) |>
  show_table()
```

## Marginal CS and AR(1) comparisons with `nlme::gls()`

These models contain no random effect. The repeated-measures covariance is
compound symmetric or AR(1), and both implementations optimize a REML
criterion.

```{r marginal-nlme-fits, eval=has_nlme}
marginal.fits <- list()
marginal.refs <- list()

for (structure in c("cs", "ar1")) {
  marginal.fits[[structure]] <- lmm(
    orthodont,
    distance ~ age + Sex,
    repeated = ~ Occasion | Subject,
    structure = structure,
    method = "REML"
  )

  correlation <- if (structure == "cs") {
    nlme::corCompSymm(form = ~ 1 | Subject)
  } else {
    nlme::corAR1(form = ~ 1 | Subject)
  }

  marginal.refs[[structure]] <- nlme::gls(
    distance ~ age + Sex,
    data = orthodont,
    correlation = correlation,
    method = "REML"
  )
}
```

```{r marginal-nlme-fixed, eval=has_nlme}
marginal.fixed <- do.call(rbind, lapply(c("cs", "ar1"), function(structure) {
  fit <- marginal.fits[[structure]]
  ref <- marginal.refs[[structure]]
  table <- comparison_table(
    term = names(fixef(fit)),
    lmmix = fixef(fit),
    reference = stats::coef(ref)
  )
  tibble::add_column(table, structure = structure, .before = 1)
}))

show_table(marginal.fixed)
```

```{r marginal-nlme-covariance, eval=has_nlme}
marginal.covariance <- do.call(
  rbind,
  lapply(c("cs", "ar1"), function(structure) {
    fit <- marginal.fits[[structure]]
    ref <- marginal.refs[[structure]]
    reference.parameters <- c(
      ref$sigma^2,
      as.numeric(stats::coef(
        ref$modelStruct$corStruct,
        unconstrained = FALSE
      ))
    )
    table <- comparison_table(
      term = c("residual.var", "residual.cor"),
      lmmix = VarCorr(fit)$estimate,
      reference = reference.parameters
    )
    tibble::add_column(table, structure = structure, .before = 1)
  })
)

show_table(marginal.covariance)

tibble::tibble(
  structure = c("cs", "ar1"),
  lmmix = vapply(
    marginal.fits,
    function(x) as.numeric(logLik(x)),
    numeric(1)
  ),
  reference = vapply(
    marginal.refs,
    function(x) as.numeric(logLik(x)),
    numeric(1)
  )
) |>
  transform(abs.diff = abs(lmmix - reference)) |>
  show_table()
```

## Marginal AR(1) comparison with `mmrm::mmrm()`

The `mmrm` comparison provides a second independent implementation of the
marginal AR(1) model. It checks the complete residual covariance matrix rather
than only the variance and correlation parameters.

```{r marginal-mmrm, eval=has_nlme && has_mmrm}
ref.mmrm <- mmrm::mmrm(
  distance ~ age + Sex,
  data = orthodont,
  covariance = mmrm::cov_struct(
    type = "ar1",
    visits = "Occasion",
    subject = "Subject"
  ),
  reml = TRUE
)

fit.mmrm <- marginal.fits[["ar1"]]

comparison_table(
  term = names(fixef(fit.mmrm)),
  lmmix = fixef(fit.mmrm),
  reference = stats::coef(ref.mmrm)
) |>
  show_table()

tibble::tibble(
  quantity = c("logLik", "residual.covariance.max.abs.diff"),
  lmmix = c(as.numeric(logLik(fit.mmrm)), NA_real_),
  reference = c(as.numeric(logLik(ref.mmrm)), NA_real_),
  abs.diff = c(
    abs(as.numeric(logLik(fit.mmrm)) - as.numeric(logLik(ref.mmrm))),
    max(abs(
      unname(fit.mmrm$covariance$residual_base) -
        unname(mmrm::component(ref.mmrm, "varcor"))
    ))
  )
) |>
  show_table()
```

## Satterthwaite comparison with `lmerTest::lmer()`

The covariance model in this section is a random intercept with independent
residuals, which is supported by both packages. The coefficient comparison
checks estimates, standard errors, and one-dimensional Satterthwaite degrees
of freedom.

```{r lmer-test-coefficients, eval=has_nlme && has_lmer_test}
ref.lmer <- lmerTest::lmer(
  distance ~ age + Sex + (1 | Subject),
  data = orthodont,
  REML = TRUE
)

lmmix.coef <- generics::tidy(fit.ri)
lmer.coef <- summary(ref.lmer)$coefficients

tibble::tibble(
  term = lmmix.coef$term,
  lmmix.estimate = lmmix.coef$estimate,
  lmerTest.estimate = lmer.coef[, "Estimate"],
  estimate.abs.diff = abs(lmmix.estimate - lmerTest.estimate),
  lmmix.std.error = lmmix.coef$std.error,
  lmerTest.std.error = lmer.coef[, "Std. Error"],
  std.error.abs.diff = abs(lmmix.std.error - lmerTest.std.error),
  lmmix.df = lmmix.coef$df,
  lmerTest.df = lmer.coef[, "df"],
  df.abs.diff = abs(lmmix.df - lmerTest.df)
) |>
  show_table()
```

The type III comparison uses the same fixed-effect terms and the same
Satterthwaite denominator-df method.

```{r lmer-test-type3, eval=has_nlme && has_lmer_test}
lmmix.type3 <- anova(fit.ri)
lmer.type3 <- anova(ref.lmer, type = 3)

tibble::tibble(
  term = lmmix.type3$term,
  lmmix.num.df = lmmix.type3$num.df,
  lmerTest.num.df = lmer.type3[, "NumDF"],
  num.df.abs.diff = abs(lmmix.num.df - lmerTest.num.df),
  lmmix.den.df = lmmix.type3$den.df,
  lmerTest.den.df = lmer.type3[, "DenDF"],
  den.df.abs.diff = abs(lmmix.den.df - lmerTest.den.df),
  lmmix.statistic = lmmix.type3$statistic,
  lmerTest.statistic = lmer.type3[, "F value"],
  statistic.abs.diff = abs(lmmix.statistic - lmerTest.statistic)
) |>
  show_table()
```

## Combined covariance structures

The included `multicentre` data reproduce the multilocation repeated-measures
example in Section 28.3 of @milliken2009. They contain 153 rows, including 28
missing responses. The complete-case models therefore use 125 observations.

No single external R function used above fits all of the following components
in one object: a random center effect, a structured within-subject residual
covariance, and the `lmmix` Satterthwaite inference layer. For this combined
case, all five residual structures are checked for optimizer convergence, a
positive-definite likelihood Hessian, and a positive-definite marginal
covariance matrix.

```{r combined-structures}
combined.fits <- lapply(
  c("id", "cs", "ar1", "toep", "un"),
  function(structure) {
    lmm(
      multicentre,
      Y ~ Drug * Time,
      random = ~ 1 | Center,
      repeated = ~ Time | Center:Drug:Subject,
      structure = structure,
      method = "REML",
      ddf = "satterthwaite"
    )
  }
)
names(combined.fits) <- c("id", "cs", "ar1", "toep", "un")

tibble::tibble(
  structure = names(combined.fits),
  convergence = vapply(
    combined.fits,
    function(x) x$convergence$code,
    integer(1)
  ),
  hessian.positive.definite = vapply(
    combined.fits,
    function(x) x$convergence$hessian_positive_definite,
    logical(1)
  ),
  min.marginal.eigenvalue = vapply(
    combined.fits,
    function(x) {
      min(eigen(
        x$covariance$v,
        symmetric = TRUE,
        only.values = TRUE
      )$values)
    },
    numeric(1)
  )
) |>
  show_table()
```

## Stored PROC MIXED comparison

The stored targets correspond to the following specification. The RANDOM and
REPEATED statements represent distinct covariance sources, and
`DDFM=SATTERTHWAITE` requests approximate denominator degrees of freedom as
documented by @sas2015.

```sas
proc mixed data=multicentre method=reml;
  class Center Drug Subject Time;
  model Y = Drug Time Drug*Time / ddfm=satterth solution;
  random Center;
  repeated Time / type=ar(1) subject=Subject(Center*Drug);
  lsmeans Drug Time / diff;
run;
```

The exact `lmmix` fit used in the comparison is:

```{r sas-fit}
sas.fit <- combined.fits[["ar1"]]
sas.fit
```

### Covariance parameters and likelihood criterion

```{r sas-covariance}
sas.covariance.target <- c(5.1737, 10.6702, 0.9351)
lmmix.covariance <- VarCorr(sas.fit)

tibble::tibble(
  parameter = paste(
    lmmix.covariance$group,
    lmmix.covariance$term,
    lmmix.covariance$component,
    sep = "."
  ),
  lmmix = lmmix.covariance$estimate,
  SAS.target = sas.covariance.target,
  abs.diff = abs(lmmix - SAS.target)
) |>
  show_table()

tibble::tibble(
  quantity = "deviance",
  lmmix = deviance(sas.fit),
  SAS.target = 490.62570532,
  abs.diff = abs(lmmix - SAS.target)
) |>
  show_table()
```

### Fixed effects

```{r sas-fixed-effects}
sas.fixed.target <- c(
  12.0524590,
  4.7647059,
  3.9411765,
  0.9210892,
  2.3922237,
  -0.1427475,
  -0.4735052,
  0.8233152,
  0.3342713
)

sas.fixed.table <- comparison_table(
  term = names(fixef(sas.fit)),
  lmmix = fixef(sas.fit),
  reference = sas.fixed.target
)
names(sas.fixed.table) <- c("term", "lmmix", "SAS.target", "abs.diff")
show_table(sas.fixed.table)
```

### Type III tests

The stored type III targets contain F statistics rounded to two decimal places
and the interaction p-value. The other two SAS p-values were not available in
the stored target set and are therefore reported as `not.stored` rather than
reconstructed.

```{r sas-type3}
lmmix.sas.type3 <- anova(sas.fit)

tibble::tibble(
  term = lmmix.sas.type3$term,
  lmmix.num.df = lmmix.sas.type3$num.df,
  lmmix.den.df = lmmix.sas.type3$den.df,
  lmmix.statistic = lmmix.sas.type3$statistic,
  SAS.statistic = c(11.43, 59.27, 1.35),
  statistic.abs.diff = abs(lmmix.statistic - SAS.statistic),
  lmmix.p.value = lmmix.sas.type3$p.value,
  SAS.p.value = c("not.stored", "not.stored", "0.2597")
) |>
  show_table()
```

### LS-means

The following table compares every stored Drug and Time marginal mean,
standard error, and denominator df.

```{r sas-lsmeans}
drug.means <- lsmeans(sas.fit, ~Drug)
time.means <- lsmeans(sas.fit, ~Time)

lmmix.means <- rbind(
  tibble::tibble(
    factor = "Drug",
    level = as.character(drug.means$Drug),
    estimate = drug.means$estimate,
    std.error = drug.means$std.error,
    df = drug.means$df
  ),
  tibble::tibble(
    factor = "Time",
    level = as.character(time.means$Time),
    estimate = time.means$estimate,
    std.error = time.means$std.error,
    df = time.means$df
  )
)

sas.mean.estimate <- c(
  13.1568, 18.1484, 17.0516,
  14.9543, 15.6700, 17.7324
)
sas.mean.std.error <- c(
  1.5290, 1.5324, 1.5327,
  1.3961, 1.3985, 1.4032
)
sas.mean.df <- c(2.98, 3.01, 3.01, 2.08, 2.09, 2.12)

tibble::tibble(
  factor = lmmix.means$factor,
  level = lmmix.means$level,
  lmmix.estimate = lmmix.means$estimate,
  SAS.estimate = sas.mean.estimate,
  estimate.abs.diff = abs(lmmix.estimate - SAS.estimate),
  lmmix.std.error = lmmix.means$std.error,
  SAS.std.error = sas.mean.std.error,
  std.error.abs.diff = abs(lmmix.std.error - SAS.std.error),
  lmmix.df = lmmix.means$df,
  SAS.df = sas.mean.df
) |>
  show_table()
```

### Pairwise LS-mean differences

```{r sas-lsmeans-pairs}
drug.pairs <- lsmeans(sas.fit, pairwise ~ Drug)$contrasts
time.pairs <- lsmeans(sas.fit, pairwise ~ Time)$contrasts

lmmix.pairs <- rbind(
  tibble::add_column(drug.pairs, factor = "Drug", .before = 1),
  tibble::add_column(time.pairs, factor = "Time", .before = 1)
)

sas.pair.estimate <- c(
  -4.9916, -3.8948, 1.0968,
  -0.7157, -2.7781, -2.0624
)
sas.pair.std.error <- c(
  1.0990, 1.0987, 1.1043,
  0.1845, 0.2714, 0.2056
)
sas.pair.df <- c(46.1, 46.1, 46.9, 68.1, 73.2, 68.6)
sas.pair.p.value <- c(
  "0.00004", "0.0009", "0.3257",
  "0.0002", "<0.0001", "<0.0001"
)

tibble::tibble(
  factor = lmmix.pairs$factor,
  contrast = lmmix.pairs$contrast,
  lmmix.estimate = lmmix.pairs$estimate,
  SAS.estimate = sas.pair.estimate,
  estimate.abs.diff = abs(lmmix.estimate - SAS.estimate),
  lmmix.std.error = lmmix.pairs$std.error,
  SAS.std.error = sas.pair.std.error,
  std.error.abs.diff = abs(lmmix.std.error - SAS.std.error),
  lmmix.df = lmmix.pairs$df,
  SAS.df = sas.pair.df,
  lmmix.p.value = lmmix.pairs$p.value,
  SAS.p.value = sas.pair.p.value
) |>
  show_table()
```

Estimates, covariance parameters, and standard errors are tested with absolute
tolerance `1e-3`. Degrees of freedom are compared at the precision stored in
the targets. Type III F statistics are compared after rounding to two decimal
places. Inequalities such as `<0.0001` are checked as inequalities, not as
invented exact p-values.

## Automated regression tests

The numerical claims in this vignette are enforced in the package test suite.

| test file | enforced comparison |
|:--|:--|
| `test-likelihood.R` | `nlme::lme()`, `nlme::gls()`, and `mmrm::mmrm()` likelihood and covariance comparisons |
| `test-inference.R` | `lmerTest::lmer()` coefficient df and type III Satterthwaite comparisons |
| `test-validation-sas.R` | every stored PROC MIXED covariance, fixed-effect, LS-mean, and difference target |
| `test-covariance.R` | positive definiteness and structure-specific covariance checks |
| `test-methods.R` | S3 outputs, `ggplot2` diagnostics, and readable printed table headings |

## Reproducibility information

The versions used to build this vignette are reported rather than assumed.

```{r validation-versions}
validation.packages <- c("lmmix", "nlme", "lmerTest", "mmrm")

tibble::tibble(
  package = validation.packages,
  version = vapply(
    validation.packages,
    function(package) {
      if (requireNamespace(package, quietly = TRUE)) {
        as.character(utils::packageVersion(package))
      } else {
        "not.installed"
      }
    },
    character(1)
  )
) |>
  show_table()

tibble::tibble(
  R.version.string = R.version.string,
  platform = R.version$platform
) |>
  show_table()
```

## Boundaries of the evidence

The validation does not cover Kenward-Roger inference because it is not
implemented. It does not validate generalized responses, multiple independent
random-effect terms, likelihood-ratio model comparison, or large-scale sparse
performance. The current combined-model evidence consists of structural tests
and stored PROC MIXED regression targets. A fresh SAS execution with an
archived log and output would be required to upgrade the SAS evidence to a new
independent run.
